---
title: "Juvenile snowshoe hare sizes: exploratory report"
author: "By: Daija Odom"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: yes
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)
library(janitor)
library(effsize)
library(broom)
library(kableExtra)
library(ggbeeswarm)
library(ggpubr)
library(here)
library(lubridate)

```

## Introduction

### 1. Overview

This report provides an exploratory overview of counts and differences in juvenile snowshoe hare (*Lepus americanus*) sizes in the Bonanza Creek Experimental Forest from 1999 - 2012. Counts were taken from 3 sampling sites (Bonanza Riparian, Bonanza Mature, and Lowland Black Spruce). Specifically, juvenile hare weights are compared by sex and site. And the relationship changes in respect to hind foot length are explored.

### 2. Data and methods

Snowshoe hare capture-recapture was conducted at 5 locales in the Tanana valley and made available by Associate Professor Knut Kielland and colleagues at the Bonanza Creek Long Term Ecological Research (LTER) Network site (Kielland et al. 2017). Following exploratory data visualization, hare weight between male and female juvenile hares are by unpaired two-sample t-tests using a significance level of ($\alpha$) of 0.05 throughout. Differences between groups are described by Cohen's d effect size and the relationship between hind foot length and hare weight is explored by simple linear regression. All analyses are in R version 4.1.1 using RStudio version 1.4.1717. 


```{r}
bonanza_hares <- read.csv(here("data", "bonanza_hares.csv")) 
```


```{r}
juvenile_hares <- bonanza_hares %>% # Create a subset of data only containing juvenile hares
  mutate(date = mdy(date)) %>% # changed the class to dates
  mutate(year = year(date)) %>%  # added a new column with only year
  filter(age == "j") %>% 
  mutate(sex = case_when(
    sex == "m" ~ "Male",
    sex == "f" ~ "Female",
  ))
  # filtering to only include juvenile hares
# class(juvenile_hares$date) # checked the class
```


```{r}
juvenile_counts <- juvenile_hares %>% # create subset of data containing counts of hares during each year
  group_by(year) %>% # grouped counts by year
  summarize(hare_count = n()) #create a column of hare count holding the number
           

mean_hare_count <- mean(juvenile_counts$hare_count, na.rm = TRUE) # call the mean of juvenile hares and skip NA values
median_hare_count <- median(juvenile_counts$hare_count, na.rm = TRUE) # call the median of juvenile hares and skip NA values
min_hare_count <- min_hare_count <- as.numeric(min(juvenile_counts$hare_count, na.rm = TRUE)) #call the minimum number hares
max_hare_count <- max_hare_count <- as.numeric(max(juvenile_counts$hare_count, na.rm = TRUE)) # call the maximum of hares

# class(juvenile_counts$hare_count)
```

### 3. Preliminary results

#### A. Number of juvenile hare trappings comparison


```{r, fig.cap= "**Figure 1.** Total number of juvenile snowshoe hare trappings (male and female) annually."}

juvenile_counts$year <- as.factor(juvenile_counts$year) # made year a factor
ggplot(data = juvenile_counts, aes(x = year, y = hare_count)) +
       geom_col(aes(fill = hare_count, 
                    color = hare_count),
                show.legend = FALSE) + # removed legend for redundancy 
  scale_color_viridis_c() + # added colorblind scale for continuous value
  scale_fill_viridis_c() + # added colorblind scale for continuous value
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +
              #scale_x_continuous(breaks = 1999:2012) +
                  labs(x = "Year",
                       y = "Hare Count",
                       title = "Annual Juvenile Snowshoe Hare Trappings")
```

The minimum number of juvenile hares trapped are `r min(juvenile_counts$hare_count, na.rm = TRUE)` and the maximum number of juvenile hares trapped are `r max(juvenile_counts$hare_count, na.rm = TRUE)`. The mean annual number of hares trapped is `r mean(juvenile_counts$hare_count, na.rm = TRUE)` and the median is `r median(juvenile_counts$hare_count, na.rm = TRUE)`. Hare counts are highest in 1999 and overall decrease in the years thereafter with the smallest count being in 2010. To standardize the juvenile hare population in future analyses I suggest monitoring vegetation and predators in each site which affect changes in the hare densities and setting up the traps on the same days every year.

#### B. Juvenile hare weights comparison 

```{r, fig.cap = "**Figure 2.** Hare weight observations (g) for juvenile female and male snowshoe hares. Maroon (female), cerulean (male), and gray (NA) points indicate individual observations for juvenile hare weights (grams). Data: Kielland et al. (2017)."}

hare_weights <- juvenile_hares %>% # create subset of data only containing... 
  select(sex, grid, weight) %>% # ...hare weight, sex, and grid
      mutate(grid = case_when( # updated facet headers
             grid == "bonbs" ~ "Lowland Black Spruce",
             grid == "bonmat" ~ "Bonanza Mature",
             grid == "bonrip" ~ "Bonanza Riparian"))
```


```{r, fig.cap = "**Figure 3.** Hare weight observations (g) for juvenile female and male snowshoe hares. Violet-red (female), sky blue (male), and gray (NA) points indicate individual observations for juvenile hare weights (grams). Data: Kielland et al. (2017)."}
ggplot(data = hare_weights, aes(x = sex, y = weight)) +
  geom_jitter(aes(color = sex)) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        legend.title = element_text(color = "black")) +
  labs(x = "Sex",
       y = "Juvenile Hare Weight (g)",
       title = "Juvenile Hare Weight by Sex in Sites",
       color = "Sex") + # changed legend title to capitalize "Sex"
  facet_wrap(~ grid)

```

This graph shows normal distribution for both female and male juvenile snowshoe hare weight. Bonanaza Riparian appears to have the most observations for both males and females. The number of observations could involve the actual population size at each site due to study efforts.  

```{r, include = FALSE}
juvenile_summary <- juvenile_hares %>% 
  group_by(sex, grid) %>% 
  summarize(mean_weight = round(mean(weight, na.rm = TRUE), 2), # Rounded decimal to 2 places
            sd_weight = round(sd(weight, na.rm = TRUE), 2),
            count = n()) %>% 
  kable(col.names = c("Sex", "Grid", "Mean (g)", "Standard Deviation (g)", "Sample Size")) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE) # Placed spaces between column headers

juvenile_summary
```


```{r, fig.cap = "**Figure 4.** Hare weight observations (g) for juvenile female and male snowshoe hares. Violet-red (female), sky blue (male), and gray (NA) points indicate individual observations for juvenile hare weights (grams). Box endpoints indicate the 25th and 75th percentile values; the gray line and black point within the box indicate the median and mean value for each species, respectively. Data: Kielland et al. (2017)."}

ggplot(data = hare_weights, aes(x = sex, y = weight)) +
  geom_beeswarm(aes(color = sex)) + # visualizes summary statistics and data
  scale_color_manual(values = c("palevioletred", "skyblue1","gray")) +
  geom_boxplot(fill = NA, width = 0.2, outlier.color = NA) +
  stat_summary(fun = mean,
               geom = "point",
               shape = 20,
               size = 4,
               color = "black",
               fill = "black") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) + # positioned the plot title in the center
  labs(x = "Sex", #labeled the x and y axis
        y = "Juvenile Hare Weight (g)",
        title = "Juvenile Hare Weights", # labeled the title
       color = "Sex") #updated the legend title to capitalize "Sex"

```
 
Visualizations show juvenile snowshoe hare weight by sex (male and female) with summary statistics (mean and median).

#### C. Juvenile weight comparison (male and female snowshoe hares)
```{r, include = FALSE}
juvenile_comparison <- juvenile_hares %>% 
  filter(is.na(sex) == FALSE) %>% # filter out rows by sex, if they are not NA include them
  group_by(sex) %>% 
  summarize(mean_weight = round(mean(weight, na.rm = TRUE), 2), # Rounded decimal to 2 places
            sd_weight = round(sd(weight, na.rm = TRUE), 2),
            count = n()) %>% 
  kable(col.names = c("Sex", "Mean (g)", "Standard Deviation (g)", "Sample Size"), # create a table of summary statistics
        caption = "<b>Table 1.</b> Descriptive Statistics (mean (g), standard deviation (g), and sample size) of juvenile hare weight. Data: Kielland et al. (2017).") %>% # bolded "Table 1."
  kable_styling(bootstrap_options = "striped", full_width = FALSE) # Placed spaces between column headers

juvenile_comparison # Call on the summary table to appear in knitted document

```


```{r, include = FALSE}
ggplot(data = juvenile_hares, aes(x = weight)) +
  geom_histogram(bins = 12) + # created a histogram and reduced the bin sizes
  facet_wrap(~ sex) #faceted by sex
```

```{r, include = FALSE}
ggplot(data = juvenile_hares, aes(sample = weight)) +
  geom_qq() + # created a q-q plot
  facet_wrap(~ sex) + # faceted by sex
  geom_qq_line() # computed the slope of the line connecting points st specified quartiles of theoretical and sample distribution
```

```{r}
hare_weight_males <- juvenile_hares %>% # storing male weights in new subset
  filter(sex == "Male") %>% # filter out male
  drop_na(weight) %>% # drop the weights with NA
  pull(weight) # pull out a vector with just the weight column

hare_weight_females <- juvenile_hares %>% # storing female weights in new subset
  filter(sex == "Female") %>% # filter out female
  drop_na(weight) %>% # drop the weights with NA
  pull(weight)# pull out a vector with the just the weight column

hares_ttest <- t.test(hare_weight_males, hare_weight_females) # performed two sample t-test on vectors
#hares_ttest # call on t-test
```


```{r}
hare_weight_cohen <- cohen.d(hare_weight_males, hare_weight_females) # computed the cohen's d effect size statistics

#hare_weight_cohen # called on cohen's d
```

There is an actual difference of 90.47 (g) in means. Mean hare weights for female juvenile snowshoe hares (855.39 g $\pm$ 292.25 g; mean $\pm$ 1 standard deviation, n = 200) were only slightly smaller than those of male juvenile snowshoe hares (945.86 $\pm$ 333.22 cm; mean $\pm$ 1 standard deviation n = 163), a difference of just over 10.05%. The difference in means is statistically significant (Welch's two-sample t-test: t(325.02) = 2.71, p = 0.007). The effect size was small (Cohen's *d* = `r hare_weight_cohen$estimate`).


#### D. Relationship between juvenile weight and hind foot length

```{r, fig.cap = "**Figure 5.** Relationship beteeen hind foot length (mm) and weight change (g) for the juvenile snowshoe hares. Points indicate individual hare measurements (female = red circles; male = green circles; and NA = blue circles). Linear model summary"}
ggplot(data = juvenile_hares, aes(x = hindft, y = weight)) + 
  geom_point(aes(color = sex, pch = sex)) +
  scale_color_manual(values = c("goldenrod2","cyan4","coral"), 
                     name = "Sex") +
  scale_shape_manual(values = c(16,17,18), 
                     name = "Sex") + # changed shape of plotted points
  theme_minimal() + # changed the theme
  theme(plot.title = element_text(hjust = 0.5)) + # positioned plot title to center
  labs(x = "Hind Foot Length(mm)",
       y = "Weight (g)",
       title = "Juvenile Hare Weights on Hind Foot Length",
       color = "Sex") +
  geom_smooth(method = "lm", color = "black")

```


```{r, include = FALSE}
hare_lm <- lm(weight ~ hindft, data = juvenile_hares) # use linear regression to compare hind foot length and weight
hare_lm # call on linear regression

hare_lm_tidy <- tidy(hare_lm) # get tidy versions of the data to call on later
hare_lm_glance <- glance(hare_lm)

summary(hare_lm) # Return the complete overview
#plot(hare_lm)# automatically create four useful visualizations to consider assumptions

hare_cor <-cor.test(juvenile_hares$hindft, juvenile_hares$weight)

hare_cor_tidy <- tidy(hare_cor)
```

As a starting point, the relationship between hare weight and hind foot length was explored across the snowshoe hares (i.e. ignoring site and sex). 

The relationship between hare weight and hind foot length were not linear (Figure 5). Simple linear regression revealed that hind foot length does significantly predict hare weight (p > 0.001, R^2^ = `r round(hare_lm_glance$r.squared,2)`) with an average slope of $\beta$ = `r round(hare_lm_tidy$estimate[2], 2)` g mm^-1^ (i.e., for each one millimeter increase in hind foot length we expect an average increase in body mass of `r round(hare_lm_tidy$estimate[2], 2)` g). Hind foot length and body mass are moderately, positively correlated (Pearson's *r* = `r round(hare_cor_tidy$estimate,2)`, p < 0.001). Diagnostic plots (not included) reveal non-normally distributions and heteroscedastic residuals. 

## Summary

The exploratory data analysis reveals the following initial findings:

- There are inconsistencies within the overall trapping observations of the data with observations missing in 2002 and 2009.
- On average, juvenile snowshoe hare females have smaller body weights than juvenile snowshoe hare males.
- The relationship between hind foot length (mm) and body mass (g) does not appear to be best representend by a simple linear regression. The p-value tells you that this is significant, but the models are not right.In simple linear regression `r round(hare_lm_glance$r.squared, 2)*100` % of the variance is based on body mass. Heteroscedastic because the spread of the residuals change about the model in a non-random way.

Next steps:

- Compare hind foot lengths statistically (one-way ANOVA), including snowshoe hare sex as a covariate.
- Multiple linear regression to understand the relationship between sex, site, body mass, and hind foot length.
- Analysis of nesting pairs to explore size correlation.

## Citations

Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14
